\TheTitle

### 2019-11-20 Paper1/appendix.tex
first corrections from Martin added

\appendix
\section{Costs of evaluating the numerical flux}
\label{app:costNumFlux}
In the following, we briefly discuss the number of operations needed when precomputing integrals versus the use of a kinetic flux for Burgers' equation. The stochastic Burgers' equation reads
\begin{align*}
\partial_t u + \partial_x \frac{u^2}{2} &= 0,\\
u(t=0,x,\xi) &= u_{IC}(x,\xi).
\end{align*}
The scalar random variable $\xi$ is uniformly distributed in the interval $[-1,1]$, hence the gPC basis functions $\bm\varphi=(\varphi_0,\cdots,\varphi_M)^T$ are the Legendre polynomials. Choosing the SG ansatz \eqref{eq:SGClosure} and testing with the gPC basis functions yields the SG moment system
\begin{align*}
\partial_t \hat u_i + \partial_x \frac12\sum_{n,m = 0}^M \hat u_n \hat u_m \langle \varphi_n\varphi_m\varphi_i \rangle = 0.
\end{align*}
Defining the matrices $\bm C_i := \langle \bm\varphi\bm\varphi^T\varphi_i\rangle\in\mathbb{R}^{N\times N}$ gives
\begin{align*}
\partial_t \bm{\hat u} + \partial_x \bm F(\bm{\hat u}) = 0
\end{align*}
with $F_i(\bm{\hat u}) = \frac12\bm{\hat u}^T\bm C_i\bm{\hat u}$. Note that $\bm{C}_i$ can be computed analytically, hence choosing a Lax-Friedrichs flux
\begin{align}\label{eq:numFluxAnalytic}
G_i^{(LF)}(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) =\frac{1}{4}\left(\bm{\hat u}_{\ell}^T \bm{C}_i \bm{\hat u}_{\ell}+\bm{\hat u}_{r}^T \bm{C}_i \bm{\hat u}_{r}\right) - \frac{\Delta x}{2\Delta t}(\bm{\hat u}_{r}-\bm{\hat u}_{\ell})_i
\end{align}
requires no integral evaluations. Recall, that the numerical flux choice made in this work gives
\begin{align}\label{eq:numericalFluxIPMBurgers}
 \bm{G}(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) = \sum_{k=1}^Q w_k g(\mathcal{U}(\bm{\hat u}_{\ell};\xi_k),\mathcal{U}(\bm{\hat u}_{r};\xi_k))\bm{\varphi}(\xi_k)f_{\Xi}(\xi_k),
\end{align}
where $\mathcal{U}$ is the SG ansatz \eqref{eq:SGClosure}. When the chosen deterministic flux $g$ is Lax-Friedrichs, the order of the polynomials inside the sum is $3M=3(N-1)$. Choosing a Gauss-Lobatto quadrature rule, $Q = \frac32 N -1$ quadrature points suffice for an exact computation of the numerical flux. Indeed, with this choice of quadrature points, the numerical fluxes \eqref{eq:numFluxAnalytic} and \eqref{eq:numericalFluxIPMBurgers} are equivalent. 
Counting the number of operations, one observes that our choice of the numerical flux \eqref{eq:numericalFluxIPMBurgers} is significantly cheaper: When computing and storing the values in a matrix $\bm A\in\mathbb{R}^{Q\times N}$ with entries $a_{ki} = \varphi_i(\xi_k)$ before running the program, the numerical flux \eqref{eq:numericalFluxIPMBurgers} can be split into two parts. First, we determine the SG solution at all quadrature points, i.e. we compute $\bm{u}^{(\ell)} := \bm A \bm{\hat u}_{\ell}$ and $\bm{u}^{(r)} := \bm A \bm{\hat u}_{r}$ which requires $O(N\cdot Q)$ operations. These solution values are then used to compute the numerical flux
\begin{align*}
G_i(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) &= \sum_{k=1}^Q w_k g(u^{(\ell)}_k,u^{(r)}_k)a_{ki}f_{\Xi}(\xi_k),
\end{align*}
which again requires $O(N\cdot Q)$ operations, i.e. the costs are $O(N^2)$. The evaluation of \eqref{eq:numFluxAnalytic} however requires $O(N^3)$ operations.

>>>>

\appendix
\section{Costs of evaluating the numerical flux}
\label{app:costNumFlux}
In the following, we briefly discuss the number of operations needed when precomputing integrals versus the use of a kinetic flux for Burgers' equation. The stochastic Burgers' equation reads
\begin{align*}
\partial_t u + \partial_x \frac{u^2}{2} &= 0,\\
u(t=0,x,\xi) &= u_{IC}(x,\xi).
\end{align*}
The scalar random variable $\xi$ is uniformly distributed in the interval $[-1,1]$, hence the gPC basis functions $\bm\varphi=(\varphi_0,\cdots,\varphi_M)^T$ are the Legendre polynomials. Choosing the SG ansatz \eqref{eq:SGClosure} and testing with the gPC basis functions yields the SG moment system
\begin{align*}
\partial_t \hat u_i + \partial_x \frac12\sum_{n,m = 0}^M \hat u_n \hat u_m \langle \varphi_n\varphi_m\varphi_i \rangle = 0.
\end{align*}
Defining the matrices $\bm C_i := \langle \bm\varphi\bm\varphi^T\varphi_i\rangle\in\mathbb{R}^{N\times N}$ gives
\begin{align*}
\partial_t \bm{\hat u} + \partial_x \bm F(\bm{\hat u}) = 0
\end{align*}
with $F_i(\bm{\hat u}) = \frac12\bm{\hat u}^T\bm C_i\bm{\hat u}$. Note that $\bm{C}_i$ can be computed analytically, hence choosing a Lax-Friedrichs flux
\begin{align}\label{eq:numFluxAnalytic}
G_i^{(LF)}(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) =\frac{1}{4}\left(\bm{\hat u}_{\ell}^T \bm{C}_i \bm{\hat u}_{\ell}+\bm{\hat u}_{r}^T \bm{C}_i \bm{\hat u}_{r}\right) - \frac{\Delta x}{2\Delta t}(\bm{\hat u}_{r}-\bm{\hat u}_{\ell})_i
\end{align}
requires no integral evaluations. Recall, that the numerical flux choice made in this work gives
\begin{align}\label{eq:numericalFluxIPMBurgers}
 \bm{G}(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) = \sum_{k=1}^Q w_k g(\mathcal{U}(\bm{\hat u}_{\ell};\xi_k),\mathcal{U}(\bm{\hat u}_{r};\xi_k))\bm{\varphi}(\xi_k)f_{\Xi}(\xi_k),
\end{align}
where $\mathcal{U}$ is the SG ansatz \eqref{eq:SGClosure}. When the chosen deterministic flux $g$ is Lax-Friedrichs, the order of the polynomials inside the sum is $3M=3(N-1)$. Choosing a Gauss-Lobatto quadrature rule, $Q = \frac32 N -1$ quadrature points suffice for an exact computation of the numerical flux. Indeed, with this choice of quadrature points, the numerical fluxes \eqref{eq:numFluxAnalytic} and \eqref{eq:numericalFluxIPMBurgers} are equivalent. 
Counting the number of operations, one observes that our choice of the numerical flux \eqref{eq:numericalFluxIPMBurgers} uses $O(N^2)$ operations whereas \eqref{eq:numFluxAnalytic} requires $O(N^3)$ operations: When computing and storing the values in a matrix $\bm A\in\mathbb{R}^{Q\times N}$ with entries $a_{ki} = \varphi_i(\xi_k)$ before running the program, the numerical flux \eqref{eq:numericalFluxIPMBurgers} can be split into two parts. First, we determine the SG solution at all quadrature points, i.e. we compute $\bm{u}^{(\ell)} := \bm A \bm{\hat u}_{\ell}$ and $\bm{u}^{(r)} := \bm A \bm{\hat u}_{r}$ which requires $O(N\cdot Q)$ operations. These solution values are then used to compute the numerical flux
\begin{align*}
G_i(\bm{\hat u}_{\ell},\bm{\hat u}_{r}) &= \sum_{k=1}^Q w_k g(u^{(\ell)}_k,u^{(r)}_k)a_{ki}f_{\Xi}(\xi_k),
\end{align*}
which again requires $O(N\cdot Q)$ operations, i.e. the costs are $O(N^2)$. The evaluation of \eqref{eq:numFluxAnalytic} however requires $O(N^3)$ operations.

### 2019-11-20 Paper1/implementation.tex
first corrections from Martin added

\section{Parallelization and Implementation}
\label{sec:parallel}
It remains to discuss the parallelization of the presented algorithms. In order to minimize the parallelization overhead, our goal is to minimize the need to send data between processors. Note that the dual problem (line 8 in Algorithm~\ref{alg:ad-IPM} and line 5 in Algorithm~\ref{alg:adosIPM}) does not have a stencil, i.e. it suffices to distribute the spatial cells between processors. In contrast to that, the finite volume update (line 14 in Algorithm~\ref{alg:ad-IPM} and line 8 in Algorithm~\ref{alg:adosIPM}) has a stencil. Hence, distributing the spatial mesh between processors will yield communication overhead since data needs to be sent whenever a stencil cell lies on a different processor. Therefore, we choose to parallelize the quadrature points, which again minimizes communication effort. As mentioned in \ref{app:costNumFlux}, we first compute the solution at stencil cells for all quadrature points. I.e. we determine $\bm u^{(j)}_k\in\mathbb{R}^p$ and the corresponding stencil cells for $k = 1,\cdots,Q$ by
\begin{align*}
\bm u^{(j-1)}_k := \bm u_{s}(\bm{\lambda}_{j-1}^T\bm{\varphi}_{\ell_1'}(\bm \xi_k)), \enskip \bm u^{(j)}_k := \bm u_{s}(\bm{\lambda}_{j}^T\bm{\varphi}_{\ell_2'}(\bm \xi_k)), \enskip \bm u^{(j+1)}_k := \bm u_{s}(\bm{\lambda}_{j+1}^T\bm{\varphi}_{\ell_3'}(\bm \xi_k)).
\end{align*}
Then, the finite volume update function \eqref{eq:adaptiveFVUpdate} can be written as
\begin{align}\label{eq:momentUpQuadrature}
\bm{c}_{\ell}^{\bm{\ell}'}&\left(\bm{\lambda}_{1},\bm{\lambda}_2,\bm{\lambda}_3\right)=\sum_{k=1}^Q w_k \left[\bm u^{(j)}_k- \frac{\Delta t}{\Delta x}\left(\bm g( \bm u^{(j)}_k,\bm u^{(j+1)}_k )- \bm g( \bm u^{(j-1)}_k,\bm u^{(j)}_k )\right)\right]\bm{\varphi}_{\ell}(\bm \xi_k)^T.
\end{align}
Instead of distributing the spatial mesh on the different processors, we now distribute the quadrature set, i.e. the sum in \eqref{eq:momentUpQuadrature} can be computed in parallel. Now after having performed the dual update, the dual variables are send to all processors. With these variables, each processor computes the solution on its portion of the quadrature set and then computed its part of the sum in \eqref{eq:momentUpQuadrature} on all spacial cells. All parts from the different processors are then added together and the full time-updated moments are distributed to all processors. From here, the dual update can again be performed. The standard IPM Algorithm~\ref{alg:IPM} and One-Shot IPM \ref{alg:osIPM} use this parallelization strategy accordingly. Again, we point out that stochastic-Galerkin is a variant of IPM, i.e. all presented techniques for IPM can also be used for SG. The SC algorithm that we use to compare intrusive with non-intrusive methods uses a given deterministic solver as a black box. Here, we distribute the quadrature set between all processors. Note that both, SC and IPM are based on the same deterministic solver, i.e. we use the same deterministic numerical flux $\bm g$. This allows a fair comparison of the different intrusive and non-intrusive techniques.

>>>>

\section{Parallelization and Implementation}
\label{sec:parallel}
It remains to discuss the parallelization of the presented algorithms. In order to minimize the parallelization overhead, our goal is to minimize the need to send data between processors. Note that the dual problem (line 8 in Algorithm~\ref{alg:ad-IPM} and line 5 in Algorithm~\ref{alg:adosIPM}) does not require communication, i.e. it suffices to distribute the spatial cells between processors. In contrast to that, the finite volume update (line 14 in Algorithm~\ref{alg:ad-IPM} and line 8 in Algorithm~\ref{alg:adosIPM}) requires communication, since values at neighboring cells need to be evaluated. Hence, distributing the spatial mesh between processors will yield communication overhead since data needs to be sent whenever a stencil cell lies on a different processor. Therefore, we choose to parallelize the quadrature points, which again minimizes communication effort. As mentioned in \ref{app:costNumFlux}, we first compute the solution at stencil cells for all quadrature points. I.e. we determine $\bm u^{(j)}_k\in\mathbb{R}^p$ and the corresponding stencil cells for $k = 1,\cdots,Q$ by
\begin{align*}
\bm u^{(j-1)}_k := \bm u_{s}(\bm{\lambda}_{j-1}^T\bm{\varphi}_{\ell_1'}(\bm \xi_k)), \enskip \bm u^{(j)}_k := \bm u_{s}(\bm{\lambda}_{j}^T\bm{\varphi}_{\ell_2'}(\bm \xi_k)), \enskip \bm u^{(j+1)}_k := \bm u_{s}(\bm{\lambda}_{j+1}^T\bm{\varphi}_{\ell_3'}(\bm \xi_k)).
\end{align*}
Then, the finite volume update function \eqref{eq:adaptiveFVUpdate} can be written as
\begin{align}\label{eq:momentUpQuadrature}
\bm{c}_{\ell}^{\bm{\ell}'}&\left(\bm{\lambda}_{1},\bm{\lambda}_2,\bm{\lambda}_3\right)=\sum_{k=1}^Q w_k \left[\bm u^{(j)}_k- \frac{\Delta t}{\Delta x}\left(\bm g( \bm u^{(j)}_k,\bm u^{(j+1)}_k )- \bm g( \bm u^{(j-1)}_k,\bm u^{(j)}_k )\right)\right]\bm{\varphi}_{\ell}(\bm \xi_k)^T.
\end{align}
Instead of distributing the spatial mesh on the different processors, we now distribute the quadrature set, i.e. the sum in \eqref{eq:momentUpQuadrature} can be computed in parallel. Now after having performed the dual update, the dual variables are send to all processors. With these variables, each processor computes the solution on its portion of the quadrature set and then computed its part of the sum in \eqref{eq:momentUpQuadrature} on all spacial cells. All parts from the different processors are then added together and the full time-updated moments are distributed to all processors. From here, the dual update can again be performed. The standard IPM Algorithm~\ref{alg:IPM} and One-Shot IPM Algorithm~\ref{alg:osIPM} use this parallelization strategy accordingly. Again, we point out that stochastic-Galerkin is a variant of IPM, i.e. all presented techniques for IPM can also be used for SG. The SC algorithm that we use to compare intrusive with non-intrusive methods uses a given deterministic solver as a black box. Here, we distribute the quadrature set between all processors. Note that both, SC and IPM are based on the same deterministic solver, i.e. we use the same deterministic numerical flux $\bm g$. This allows a fair comparison of the different intrusive and non-intrusive techniques.

### 2019-11-20 Paper1/introduction.tex
changes Paper Jonas + questions to Martin
first corrections from Martin added

Commonly, methods to quantify uncertainties use a polynomial chaos (PC) expansion \cite{wiener1938homogeneous,xiu2002wiener} to represent the solution, i.e. the uncertain space is spanned by polynomial basis functions. The remaining task is then to determine adequate expansion coefficients, often called moments or PC coefficients. Numerical methods for approximating these coefficients can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule. Commonly, SC uses sparse grids, since they posses a reduced number of collocation points for multi-dimensional problems. Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations decouple and the workload can easily be distributed across several processors.

The main idea of intrusive methods is to derive a system of equations describing the time evolution of the moments which can then be solved with a deterministic numerical scheme. A popular approach to describe the moment system is the stochastic-Galerkin (SG) method \cite{ghanem2003stochastic}, which chooses a polynomial basis ansatz of the solution and performs a Galerkin projection to derive a closed system of equations. One significant drawback of SG is, that its moment system is not necessarily hyperbolic \cite{poette2009uncertainty}. A generalization of stochastic-Galerkin, which ensures hyperbolicity is the Intrusive Polynomial Moment (IPM) method \cite{poette2009uncertainty}. Instead of performing the PC expansion on the solution, the IPM method represents the entropy variables with polynomials. Besides yielding a hyperbolic moment system, the IPM method has several advantages: Choosing a quadratic entropy yields the stochastic-Galerkin moment system, i.e. IPM generalizes different intrusive methods. Furthermore, at least for scalar problems, IPM is significantly less oscillatory compared to SG \cite{kusch2017maximum}. Also, as discussed in \cite{poette2009uncertainty}, when choosing a physically correct entropy of the deterministic problem, the IPM solution dissipates the expectation value of the entropy, i.e. the IPM method yields a physically correct entropy solution. Unfortunately, the desirable properties of IPM come along with significantly increased numerical costs, since IPM requires the repeated computation of the entropic expansion coefficients from the moment vector, which involves solving a convex optimization problem. However, IPM and minimal entropy methods in general are well suited for modern HPC architectures.

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through either the spatial numerical method \cite{offner2017stability} or by filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{wan2006multi,durrwachter2018hyperbolicity}. Discontinuous structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. In addition to that, the number of moments needed to obtain a certain order with intrusive methods is asymptotically smaller than the number of quadrature points for SC. An additional downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Consequently, collocation methods typically require a higher number of unknowns than intrusive methods to reach a given accuracy \cite{xiu2009fast,alekseev2011estimation}. Therefore, one aim should be to accelerate intrusive methods, since they can potentially outperform non-intrusive methods in complex and high-dimensional settings. \\

@@

The core idea of the stochastic-Collocation method is to compute the moments in the gPC expansion with a quadrature rule. Given a set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Hence, the moments can be computed by running a given deterministic solver for the original problem at each quadrature point. To reduce numerical costs in multi-dimensional settings, SC commonly uses sparse grids as quadrature rule: While tensorized quadrature sets require $O(M^s)$ quadrature points to integrate polynomials of total degree $M$ exactly, sparse grids only require $O(M(\log_2(M)^{s-1}))$ quadrature points.

>>>>

One strategy to represent the solution's dependence on uncertainties is to use a polynomial chaos (PC) expansion \cite{wiener1938homogeneous,xiu2002wiener}, i.e. the uncertain space is spanned by polynomial basis functions. The remaining task is then to determine adequate expansion coefficients, often called moments or PC coefficients. Numerical methods for approximating these coefficients can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule. Commonly, SC uses sparse grids, since they posses a reduced number of collocation points for multi-dimensional problems. Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations decouple and the workload can easily be distributed across several processors.

The main idea of intrusive methods is to derive a system of equations describing the time evolution of the moments which can then be solved with a deterministic numerical scheme. A popular approach to describe the moment system is the stochastic-Galerkin (SG) method \cite{ghanem2003stochastic}, which chooses a polynomial basis ansatz of the solution and performs a Galerkin projection to derive a closed system of equations. One significant drawback of SG is, that its moment system is not necessarily hyperbolic \cite{poette2009uncertainty}. A generalization of stochastic-Galerkin, which ensures hyperbolicity is the Intrusive Polynomial Moment (IPM) method \cite{poette2009uncertainty}. Instead of performing the PC expansion on the solution, the IPM method represents the entropy variables with polynomials. Besides yielding a hyperbolic moment system, the IPM method has several advantages: Choosing a quadratic entropy yields the stochastic-Galerkin moment system, i.e. IPM generalizes different intrusive methods. Furthermore, at least for scalar problems, IPM is significantly less oscillatory compared to SG \cite{kusch2017maximum}. Also, as discussed in \cite{poette2009uncertainty}, when choosing a physically correct entropy of the deterministic problem, the IPM solution dissipates the expectation value of the entropy, i.e. the IPM method yields a physically correct entropy solution. Unfortunately, the desirable properties of IPM come along with significantly increased numerical costs, since IPM requires the repeated computation of the entropic expansion coefficients from the moment vector, which involves solving a convex optimization problem. However, IPM and minimal entropy methods in general are well suited for modern HPC architectures \cite{garrett2015optimization}.

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through either the spatial numerical method \cite{offner2017stability} or by filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{wan2006multi,durrwachter2018hyperbolicity}. Discontinuous structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. In addition to that, the number of moments needed to obtain a certain order with intrusive methods is smaller than the number of quadrature points for SC. An additional downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Consequently, collocation methods typically require a higher number of unknowns than intrusive methods to reach a given accuracy \cite{xiu2009fast,alekseev2011estimation}. Therefore, one aim should be to accelerate intrusive methods, since they can potentially outperform non-intrusive methods in complex and high-dimensional settings. \\

@@

The core idea of the stochastic-Collocation method is to compute the moments in the gPC expansion with a quadrature rule. Given a set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Hence, the moments can be computed by running a given deterministic solver for the original problem at each quadrature point. To reduce numerical costs in multi-dimensional settings, SC commonly uses sparse grids as quadrature rule: While tensorized quadrature sets require $O(M^s)$ quadrature points to integrate polynomials of maximal degree $M$ exactly, sparse grids are designed to integrate polynomials of total degree $M$, for which they only require $O(M(\log_2(M)^{s-1}))$ quadrature points.

### 2019-11-20 Paper1/main.tex
changes Paper Jonas + questions to Martin
first corrections from Martin added

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{Intrusive acceleration strategies for Uncertainty Quantification}

@@

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. Non-intrusive methods allow the usage of a given deterministic solver in a black-box manner, while being embarrassingly parallel. Intrusive methods on the other hand do not suffer from aliasing effects and provide a more general framework for adaptive refinement techniques. Since the moment system solved by intrusive methods is not necessarily hyperbolic, the Intrusive Polynomial Moment (IPM) method has been introduced. IPM maintains hyperbolicity, but comes at the cost of solving an additional problem in every spatial cell and every time step.
In this work, we propose acceleration techniques for intrusive methods. When solving steady problems, the numerical costs arising from repeatedly solving the IPM optimization problem can be reduced by borrowing ideas from shape optimization. Integrating the iteration from the numerical treatment of the optimization problem into the moment update reduces numerical costs, while still guaranteeing local convergence. Additionally, we propose an adaptive implementation of the IPM method, which further reduces numerical costs. We demonstrate the effectiveness of the proposed strategies for multi-dimensional uncertainty spaces by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment with non-intrusive methods.
\end{abstract}

>>>>

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{Intrusive acceleration strategies for Uncertainty Quantification for hyperbolic systems of conservation laws}

@@

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. Non-intrusive methods allow the usage of a given deterministic solver in a black-box manner, while being embarrassingly parallel. Intrusive methods on the other hand do not suffer from aliasing effects and provide a more general framework for adaptive refinement techniques. Since the moment system solved by intrusive methods is not necessarily hyperbolic, the Intrusive Polynomial Moment (IPM) method has been introduced. IPM maintains hyperbolicity, but comes at the cost of solving an additional problem in every spatial cell and every time step.
In this work, we propose acceleration techniques for intrusive methods. When solving steady problems, the numerical costs arising from repeatedly solving the IPM optimization problem can be reduced by borrowing ideas from PDE-constrained optimization. Integrating the iteration from the numerical treatment of the optimization problem into the moment update reduces numerical costs, while still guaranteeing local convergence. Additionally, we propose an adaptive implementation of the IPM method, which further reduces numerical costs. We demonstrate the effectiveness of the proposed strategies for multi-dimensional uncertainty spaces by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment with non-intrusive methods. \frama{Sollen die Erkenntnisse vom Vergleich mit SC schon hier erw\"ahnt werden?}
\end{abstract}
